---
title: "Merced County — English Learner Trends (2020–21 to 2025–26)"
author: "Jorge Reyes-Ornelas"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 11, fig.height = 6)
library(tidyverse)
library(scales)
library(knitr)
library(tidytext)
```

# Overview

This report examines English Learner (EL) and Fluent English Proficient (FEP)
enrollment trends across five Merced County cities from 2020–21 through 2025–26.
CDE enrollment data is captured via the CALPADS census at the start of each school
year, so the 2025–26 figures reflect official counts from the August/September 2025
census and represent the full enrollment snapshot for that school year.
The period spans one of the most disruptive eras in recent public education history:
a global pandemic, remote learning, economic upheaval, and a sharp shift in federal
immigration enforcement. These are not just numbers — they reflect the conditions
of real families navigating schools, labor markets, and immigration systems
simultaneously.

Merced County is predominantly agricultural. The cities in this analysis sit at
the intersection of Central Valley farmworker communities and a growing institutional
presence anchored by the University of California, Merced. EL enrollment here is
not an abstract policy metric — it is a direct indicator of community composition,
migration patterns, and how responsive local schools are to linguistically diverse
populations.

| City | District(s) |
|------|-------------|
| Merced | Merced City Elementary, Merced Union High |
| Los Banos | Los Banos Unified |
| Atwater | Atwater Elementary |
| Livingston | Livingston Union |
| Dos Palos | Dos Palos Oro Loma Joint Unified |

---

# 1. Load & Stack All Years

```{r load}
years <- 2021:2026

sy_label <- c(
  "2021" = "2020–21",
  "2022" = "2021–22",
  "2023" = "2022–23",
  "2024" = "2023–24",
  "2025" = "2024–25",
  "2026" = "2025–26"
)

load_year <- function(prefix, year) {
  path <- paste0("data/", prefix, "_", year, ".txt")
  read_tsv(path, show_col_types = FALSE) |>
    mutate(file_year = as.character(year),
           school_year = sy_label[as.character(year)])
}

el_all  <- map_dfr(years, ~ load_year("fileselsch",  .x))
fep_all <- map_dfr(years, ~ load_year("filesfepsch", .x))
```

---

# 2. Filter to 5 Merced County Cities

```{r filter}
target_districts <- c(
  "Merced City Elementary",
  "Merced Union High",
  "Los Banos Unified",
  "Atwater Elementary",
  "Livingston Union",
  "Dos Palos Oro Loma Joint Unified"
)

city_map <- c(
  "Merced City Elementary"           = "Merced",
  "Merced Union High"                = "Merced",
  "Los Banos Unified"                = "Los Banos",
  "Atwater Elementary"               = "Atwater",
  "Livingston Union"                 = "Livingston",
  "Dos Palos Oro Loma Joint Unified" = "Dos Palos"
)

# 2025–26 reflects the CALPADS census taken at the start of the 2025–26 school year
el_merced <- el_all |>
  filter(COUNTY == "Merced", DISTRICT %in% target_districts) |>
  mutate(city = city_map[DISTRICT],
         school_year = factor(school_year, levels = unname(sy_label)))

fep_merced <- fep_all |>
  filter(COUNTY == "Merced", DISTRICT %in% target_districts) |>
  mutate(city = city_map[DISTRICT],
         school_year = factor(school_year, levels = unname(sy_label)))
```

---

# 3. Sanity Check — Schools per City per Year

```{r sanity}
el_merced |>
  distinct(school_year, city, SCHOOL) |>
  count(school_year, city) |>
  pivot_wider(names_from = city, values_from = n) |>
  kable(caption = "Distinct schools per city per school year (EL file)")
```

---

# 4. Total EL Enrollment — City Trends

```{r el-trend}
el_city_yr <- el_merced |>
  group_by(school_year, city) |>
  summarise(total_el = sum(TOTAL_EL, na.rm = TRUE), .groups = "drop")

ggplot(el_city_yr, aes(x = school_year, y = total_el, color = city, group = city)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Total English Learner Enrollment by City",
       subtitle = "Merced County — 2020–21 to 2024–25",
       x = "School Year", y = "EL Students", color = "City") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

**What this tells us:** The trend lines here reveal more than enrollment counts —
they map the presence and stability of immigrant families in each community over
time. A sustained decline in EL enrollment does not mean fewer children exist in
these cities; it may mean fewer are showing up to school. Merced, as the county
seat and largest district, carries the most EL students by volume, but the
trajectory matters as much as the magnitude. Smaller cities like Dos Palos and
Livingston, whose economies are deeply tied to agricultural labor, are more
sensitive to disruptions in migrant family stability.

```{r el-trend-table}
el_city_yr |>
  pivot_wider(names_from = city, values_from = total_el) |>
  kable(caption = "Total EL students per city per school year")
```

---

# 5. Total FEP Enrollment — City Trends

```{r fep-trend}
fep_city_yr <- fep_merced |>
  group_by(school_year, city) |>
  summarise(total_fep = sum(TOTAL, na.rm = TRUE), .groups = "drop")

ggplot(fep_city_yr, aes(x = school_year, y = total_fep, color = city, group = city)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Total Fluent English Proficient (FEP) Enrollment by City",
       subtitle = "Merced County — 2020–21 to 2024–25",
       x = "School Year", y = "FEP Students", color = "City") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

**What this tells us:** FEP students are former EL students who have been
reclassified as English proficient. A rising FEP count alongside stable or declining
EL enrollment could indicate that the reclassification pipeline is working — students
are moving through and gaining proficiency. However, FEP growth can also reflect
natural demographic aging: children of immigrants who arrived years ago and have
since acquired English. Either way, the FEP trend is an indirect indicator of
integration and language acquisition at the community level.

---

# 6. EL vs FEP — Side-by-Side by City

```{r el-fep-combined}
combined <- el_city_yr |>
  rename(count = total_el) |>
  mutate(type = "EL") |>
  bind_rows(
    fep_city_yr |>
      rename(count = total_fep) |>
      mutate(type = "FEP")
  )

ggplot(combined, aes(x = school_year, y = count, fill = type)) +
  geom_col(position = "dodge") +
  facet_wrap(~ city, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("EL" = "#2c7bb6", "FEP" = "#1a9641")) +
  labs(title = "EL vs FEP Enrollment by City",
       subtitle = "Merced County — 2020–21 to 2024–25",
       x = "School Year", y = "Students", fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold"))
```

**What this tells us:** The ratio between EL and FEP enrollment within each city
describes the integration pipeline. Cities where FEP consistently outnumbers EL
suggest a well-established, longer-tenured immigrant community — families who have
been in the area long enough for children to gain English proficiency. Cities where
EL still significantly outnumbers FEP may indicate a more recently arrived or
continuously renewing migrant population, where new arrivals regularly replenish
the EL count before families have time to stabilize.

---

# 7. Year-over-Year EL Change (%)

```{r el-yoy}
el_city_yr |>
  arrange(city, school_year) |>
  group_by(city) |>
  mutate(
    prev_el    = lag(total_el),
    pct_change = (total_el - prev_el) / prev_el
  ) |>
  filter(!is.na(pct_change)) |>
  ggplot(aes(x = school_year, y = pct_change, fill = pct_change > 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  facet_wrap(~ city) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_fill_manual(values = c("TRUE" = "#1a9641", "FALSE" = "#d7191c"),
                    guide = "none") +
  labs(title = "Year-over-Year EL Enrollment Change",
       subtitle = "Green = increase  |  Red = decrease",
       x = "School Year", y = "% Change") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold"))
```

**What this tells us:** The year-over-year view isolates momentum — where enrollment
is accelerating or losing ground. Red bars in the 2021–22 and 2022–23 window
reflect the compounding effects of COVID: school closures disrupted in-person
enrollment counts, remote learning was disproportionately harder for EL students
without reliable internet or English-speaking adults at home, and some families
returned to their countries of origin or relocated to other states. Any red bar
appearing in 2024–25 carries a different sociological weight — see Section 8 for
that analysis.

---

# 8. Immigration Policy Climate & Enrollment Decline

The year-over-year charts show enrollment fluctuations, but the *cause* of a
decline matters as much as the fact of it. This section situates the enrollment
data within the federal immigration policy timeline.

**Policy context:**

| Period | Context |
|--------|---------|
| 2020–21 | COVID-19 school closures; reduced border crossings; pandemic disruption |
| 2021–22 | Biden administration; some immigration processing resumes; enrollment partially recovers |
| 2022–23 | Title 42 lifted; increased migration; enrollment stabilizes or grows |
| 2023–24 | Bipartisan border legislation fails; enforcement rhetoric intensifies |
| 2024–25 | Trump administration returns (January 2025); mass deportation operations begin; ICE activity escalates nationally |
| 2025–26 | First full enrollment census under escalated enforcement; chilling effect on school enrollment most visible here — families counted at the start of the year reflect conditions after months of heightened ICE activity |

The mechanism connecting federal enforcement to local school enrollment is called
the **chilling effect**: when immigration enforcement becomes more visible or
aggressive, families — including those with legal status — reduce participation in
public institutions out of fear. Mixed-status households, where some members are
citizens or legal residents and others are undocumented, face the highest exposure.
Research following the 287(g) agreements and ICE workplace raids of the 2010s
consistently documented drops in school attendance and enrollment in counties with
high undocumented immigrant populations, even among U.S.-born children.

Merced County is structurally vulnerable to this effect. Agricultural labor —
the economic backbone of Livingston, Dos Palos, and Los Banos — has historically
employed a significant undocumented workforce. These are not anonymous statistics:
they are families in the same school districts, attending the same schools that
generate the EL counts in this dataset.

```{r policy-window}
# Highlight the 2024-25 year to isolate the current enforcement period
el_city_yr |>
  mutate(enforcement_era = if_else(school_year %in% c("2024–25", "2025–26"),
                                   "Escalated enforcement era",
                                   "Prior years")) |>
  ggplot(aes(x = school_year, y = total_el, group = city)) +
  geom_line(aes(color = city), linewidth = 1.2) +
  geom_point(aes(color = city, size = enforcement_era)) +
  scale_size_manual(values = c("Escalated enforcement era" = 5, "Prior years" = 2.5)) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = comma) +
  annotate("rect",
           xmin = 4.5, xmax = 6.5,
           ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "#d7191c") +
  annotate("text", x = 5.5, y = Inf, vjust = 2,
           label = "Trump admin\nJan 2025 →",
           size = 3.2, color = "#d7191c", fontface = "bold") +
  labs(title = "EL Enrollment with Federal Enforcement Context",
       subtitle = "Shaded region = escalated immigration enforcement era (2024–25 & 2025–26)",
       x = "School Year", y = "EL Students",
       color = "City", size = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

**Interpretation:** A drop in EL enrollment during 2024–25, particularly in
agricultural cities like Livingston and Dos Palos, is consistent with the documented
chilling effect of increased enforcement. These communities have the highest
structural exposure: high proportions of agricultural workers, mixed-status
households, and limited institutional buffer from legal aid or advocacy
organizations compared to urban centers like Fresno or Sacramento.

It is important to state what this data cannot prove: we cannot confirm from
enrollment records alone that fear of deportation is the cause of a decline.
Families do not report their reason for withdrawing a child from school. However,
the pattern — a drop concentrated in cities tied to agricultural labor, coinciding
with a documented escalation in federal enforcement — is sociologically coherent
and should not be dismissed as coincidence.

---

# 9. Top Languages per City (Latest Year — 2025–26)

```{r top-langs}
el_merced |>
  filter(school_year == "2025–26") |>
  group_by(city, LANGUAGE) |>
  summarise(total = sum(TOTAL_EL, na.rm = TRUE), .groups = "drop") |>
  group_by(city) |>
  slice_max(total, n = 5) |>
  ggplot(aes(x = reorder_within(LANGUAGE, total, city), y = total, fill = city)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ city, scales = "free") +
  tidytext::scale_x_reordered() +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Top 5 EL Languages per City (2024–25)",
       x = NULL, y = "EL Students") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

**What this tells us:** Language composition is a demographic fingerprint. Spanish
dominance across all five cities reflects the deep-rooted Latino presence in the
Central Valley — families tied to agricultural labor that dates back generations,
as well as more recent arrivals. The presence of other languages — Hmong in Merced
city, Punjabi in communities connected to Sikh agricultural laborers, or other
Southeast Asian languages — maps the diversity of migration streams that have
settled in this region over decades. A shift in these language rankings over time
would signal a real demographic change, not just an administrative artifact.

---

# 10. Grade Distribution — EL by City (Latest Year)

```{r grade-city}
grade_cols <- c("GR_TK","GR_KN","GR_1","GR_2","GR_3","GR_4","GR_5",
                "GR_6","GR_7","GR_8","GR_9","GR_10","GR_11","GR_12","UNGR")

el_merced |>
  filter(school_year == "2025–26") |>
  group_by(city) |>
  summarise(across(all_of(grade_cols), ~ sum(., na.rm = TRUE)), .groups = "drop") |>
  pivot_longer(-city, names_to = "grade", values_to = "count") |>
  mutate(grade = factor(grade, levels = grade_cols)) |>
  ggplot(aes(x = grade, y = count, fill = city)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "EL Students by Grade Level per City (2025–26)",
       x = "Grade", y = "Students", fill = "City") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**What this tells us:** The grade distribution reveals two distinct populations
within the EL category. High EL counts in TK through 3rd grade indicate newly
arrived children or children of recent immigrants entering the system at the
beginning. High EL counts in 6th grade and above — particularly in high school —
are a signal of Long-Term English Learners (LTELs): students who have been in the
system for six or more years without reclassifying. LTELs represent a significant
equity concern because they are often concentrated in lower-track courses, have
limited access to grade-level content, and face compounding academic disadvantage
the longer they remain in EL status. If upper-grade EL counts are substantial,
that is not a success story — it is a structural failure in the reclassification
pipeline.

---

# 11. Spanish vs All Other Languages — EL Trend

```{r spanish-trend}
el_merced |>
  mutate(lang_group = if_else(LC == "spa", "Spanish", "All Other")) |>
  group_by(school_year, city, lang_group) |>
  summarise(total = sum(TOTAL_EL, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = school_year, y = total, color = lang_group, group = lang_group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  facet_wrap(~ city, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Spanish" = "#d7191c", "All Other" = "#2c7bb6")) +
  labs(title = "Spanish vs All Other EL Languages by City",
       subtitle = "2020–21 to 2024–25",
       x = "School Year", y = "EL Students", color = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold"))
```

**What this tells us:** Spanish-speaking EL students overwhelmingly dominate in
every city in this analysis, which reflects the demographic reality of the Central
Valley. The gap between Spanish and all other languages is not just a linguistic
fact — it is a structural one. School programs, bilingual aides, translated
materials, and community outreach are all calibrated primarily around Spanish.
This is appropriate given the numbers, but it also means that non-Spanish EL
students in Merced County face a compounded disadvantage: they navigate a school
system designed for English proficiency that has limited capacity to support
languages outside of the two most common. If the "All Other" line drops in
2024–25, that may reflect displacement of non-Spanish immigrant communities —
particularly Southeast Asian families in Merced — due to enforcement pressure or
economic hardship.

---

# 12. School-Level Detail (Latest Year)

```{r school-detail}
el_merced |>
  filter(school_year == "2024–25") |>
  group_by(city, DISTRICT, SCHOOL) |>
  summarise(total_el = sum(TOTAL_EL, na.rm = TRUE), .groups = "drop") |>
  arrange(city, desc(total_el)) |>
  kable(caption = "EL students by school — 2025–26")
```

---

# 13. Key Findings

**1. Enrollment volatility maps to policy disruption, not just demographics.**
Drops in EL enrollment during 2020–21 and 2024–25 align with the two most
significant periods of external pressure on immigrant families: the COVID-19
pandemic and escalated federal immigration enforcement under the returning Trump
administration. These are not independent statistical fluctuations — they are
legible patterns when viewed through a sociological lens.

**2. Agricultural cities are the most exposed.**
Livingston, Dos Palos, and Los Banos have economies and community compositions
that make their EL populations more sensitive to immigration enforcement than
Merced city. When enforcement intensifies, the chilling effect hits hardest where
economic precarity and undocumented status intersect — which is precisely the
profile of Central Valley agricultural communities.

**3. FEP trends suggest a stabilizing long-term population alongside declining new arrivals.**
If FEP enrollment holds steady or grows while EL declines, it indicates that
established immigrant families continue to move through the integration pipeline,
but fewer new arrivals are entering the system to replenish the EL count. This
is a demographic aging pattern — not a success story, but a sign of reduced
immigration inflow.

**4. Long-Term EL risk is visible in the grade distribution.**
Substantial EL counts in grades 6–12 suggest that a significant portion of EL
students are not being reclassified on time. This is an equity issue independent
of immigration enforcement — it reflects resource allocation, program quality,
and the structural barriers EL students face in accessing grade-level content.

**5. Reclassification data ends in 2020.**
The CDE stopped publishing school-level reclassification files after 2019–20,
which removes the most direct measure of EL program effectiveness from this
analysis. EL and FEP enrollment trends are the primary longitudinal indicators
available for the post-COVID period.

---

# 14. Note on Reclassification Data

> **Data limitation:** The California Department of Education stopped publishing
> school-level reclassification files after the 2019–20 school year. The single
> snapshot included in this project (`filesreclass.txt`) represents the last
> available release. Reclassification trend analysis for 2020–21 onward is not
> possible from CDE public data. EL and FEP enrollment trends (sections 4–11)
> are the primary longitudinal indicators available for this period.

---

# Session Info

```{r session}
sessionInfo()
```
