---
title: "Merced County Housing Time Series"
name: "Jorge RO"
date: 05/11/26
output: github_document
editor_options:
  markdown:
    wrap: 72
---

# Forecasting Merced County Home Values

A Time Series Analysis Using SARIMA and ETS

## Project Overview

Where are Merced County home values headed? This project applies
classical time series methods — SARIMA and ETS — to 25 years of monthly
Zillow ZHVI data (2001–2026) for the five largest cities in Merced
County. The goal is to generate defensible 12-month forecasts and
surface the structural dynamics (trend, seasonality, volatility) that
drive local home values.

This is Part 4 of the Merced County housing series, extending the
exploratory and rental analyses with predictive modeling.

**Data:** Zillow Home Value Index (ZHVI), monthly, 2001–2026.
**Cities:** Merced, Los Banos, Atwater, Livingston, Hilmar.

> **Note on pre-processing:** The source Zillow file (`city_homevalue_zillow.csv`)
> covers every US city across 300+ monthly columns — 88 MB total. Loading
> and filtering that in R on every knit would be slow and memory-heavy.
> A Python script pre-processed it to a 1,514-row long-format CSV
> containing only the five Merced County cities, which is what this
> notebook reads.

---

## I. Environment Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(forecast)
library(tseries)
library(ggplot2)
library(scales)
library(gridExtra)

# Colors consistent with housing series
CITY_COLORS <- c(
  "Merced"     = "#1f77b4",
  "Los Banos"  = "#9467bd",
  "Atwater"    = "#d62728",
  "Livingston" = "#2ca02c",
  "Hilmar"     = "#ff7f0e"
)

df <- read_csv("merced_monthly_home_values.csv") %>%
  mutate(date = as.Date(date))

cat("Rows:", nrow(df), "\n")
cat("Date range:", format(min(df$date)), "to", format(max(df$date)), "\n")
cat("Cities:", paste(unique(df$RegionName), collapse = ", "), "\n")
```

---

## II. Historical Trends

### 25 Years of Home Value Movement

```{r historical_trends, fig.width=12, fig.height=6}
ggplot(df, aes(x = date, y = home_value, color = RegionName)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = CITY_COLORS) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  annotate("rect", xmin = as.Date("2007-01-01"), xmax = as.Date("2012-01-01"),
           ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "red") +
  annotate("rect", xmin = as.Date("2020-03-01"), xmax = as.Date("2020-09-01"),
           ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "gray40") +
  annotate("text", x = as.Date("2009-06-01"), y = 550000,
           label = "GFC", size = 3, color = "red") +
  annotate("text", x = as.Date("2020-06-01"), y = 550000,
           label = "COVID", size = 3, color = "gray40") +
  labs(title = "Merced County Home Values: 2001–2026",
       subtitle = "Monthly ZHVI | Shaded = major market disruptions",
       x = NULL, y = "Home Value", color = "City") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### Year-over-Year Growth Rate

```{r yoy_growth, fig.width=12, fig.height=5}
df %>%
  arrange(RegionName, date) %>%
  group_by(RegionName) %>%
  mutate(yoy_change = (home_value / lag(home_value, 12) - 1) * 100) %>%
  filter(!is.na(yoy_change)) %>%
  ggplot(aes(x = date, y = yoy_change, color = RegionName)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = CITY_COLORS) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  labs(title = "Year-over-Year Home Value Growth",
       subtitle = "Peak growth during 2021–2022 post-COVID boom",
       x = NULL, y = "YoY Change (%)", color = "City") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## III. Stationarity Testing

Time series models require a stationary series (constant mean and
variance). The Augmented Dickey-Fuller (ADF) test checks whether a unit
root is present — a unit root means the series is non-stationary and
must be differenced before modeling.

**H₀:** Series has a unit root (non-stationary)
**H₁:** Series is stationary

```{r stationarity}
cities <- c("Merced", "Los Banos", "Atwater", "Livingston", "Hilmar")

adf_results <- map_dfr(cities, function(city) {
  vals <- df %>% filter(RegionName == city) %>% pull(home_value)
  test <- adf.test(vals)
  tibble(
    City        = city,
    ADF_stat    = round(test$statistic, 3),
    p_value     = round(test$p.value, 4),
    Stationary  = ifelse(test$p.value < 0.05, "Yes", "No — needs differencing")
  )
})

print(adf_results)
```

### After First Differencing

```{r stationarity_diff}
adf_diff_results <- map_dfr(cities, function(city) {
  vals <- df %>% filter(RegionName == city) %>% pull(home_value)
  diff_vals <- diff(vals)
  test <- adf.test(diff_vals)
  tibble(
    City       = city,
    ADF_stat   = round(test$statistic, 3),
    p_value    = round(test$p.value, 4),
    Stationary = ifelse(test$p.value < 0.05, "Yes", "No")
  )
})

print(adf_diff_results)
```

---

## IV. Decomposition — Trend, Seasonality, Residual

STL decomposition separates each city's series into its three
structural components. This reveals how much of price movement is
driven by long-run trend vs. repeating seasonal patterns vs. irregular
shocks.

```{r decomposition, fig.width=12, fig.height=8}
# Decompose Merced as the primary city
merced_ts <- df %>%
  filter(RegionName == "Merced") %>%
  arrange(date) %>%
  pull(home_value) %>%
  ts(start = c(2001, 1), frequency = 12)

merced_stl <- stl(merced_ts, s.window = "periodic")

autoplot(merced_stl) +
  labs(title = "STL Decomposition — Merced Home Values",
       subtitle = "Trend dominates; seasonal amplitude is modest (~$5–10K)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

```{r seasonality_plot, fig.width=10, fig.height=5}
# Seasonal pattern by month
df %>%
  filter(RegionName == "Merced") %>%
  mutate(month = lubridate::month(date, label = TRUE),
         year  = lubridate::year(date)) %>%
  group_by(month) %>%
  summarise(avg_value = mean(home_value), .groups = "drop") %>%
  ggplot(aes(x = month, y = avg_value, group = 1)) +
  geom_line(color = "#1f77b4", linewidth = 1.2) +
  geom_point(color = "#1f77b4", size = 3) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Average Home Value by Month — Merced",
       subtitle = "Seasonal peak in summer, trough in winter",
       x = "Month", y = "Avg Home Value") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## V. SARIMA Modeling

### How SARIMA Works

SARIMA(**p,d,q**)(**P,D,Q**)[12] has two layers:

- **Non-seasonal** (p, d, q): AR order, differencing to remove trend,
  MA order for residual autocorrelation
- **Seasonal** (P, D, Q)[12]: Same three components but operating at
  the 12-month lag — capturing year-over-year patterns

`auto.arima()` searches the parameter space using AIC to find the
best-fitting combination automatically.

```{r sarima_fit, message=FALSE, warning=FALSE}
build_ts <- function(city_name) {
  df %>%
    filter(RegionName == city_name) %>%
    arrange(date) %>%
    pull(home_value) %>%
    ts(start = c(2001, 1), frequency = 12)
}

sarima_models <- map(setNames(cities, cities), function(city) {
  ts_data <- build_ts(city)
  auto.arima(ts_data, seasonal = TRUE, stepwise = FALSE,
             approximation = FALSE, trace = FALSE)
})

# Print model orders
map_dfr(cities, function(city) {
  m <- sarima_models[[city]]
  tibble(
    City  = city,
    Order = paste0("ARIMA(", paste(m$arma[c(1,6,2)], collapse=","), ")",
                   "(", paste(m$arma[c(3,7,4)], collapse=","), ")[12]"),
    AIC   = round(m$aic, 1),
    BIC   = round(m$bic, 1)
  )
}) %>% print()
```

### SARIMA Forecasts — All Cities

```{r sarima_forecast, fig.width=12, fig.height=8, message=FALSE, warning=FALSE}
h <- 12  # 12-month forecast horizon

sarima_forecasts <- map(setNames(cities, cities), function(city) {
  forecast(sarima_models[[city]], h = h)
})

forecast_plots <- map(cities, function(city) {
  fc <- sarima_forecasts[[city]]
  color <- CITY_COLORS[city]

  autoplot(fc) +
    scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
    labs(title = city,
         x = NULL, y = "Home Value") +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(face = "bold", color = color))
})

do.call(grid.arrange, c(forecast_plots, ncol = 2))
```

---

## VI. ETS Modeling

ETS (Error, Trend, Seasonality) uses exponentially weighted averages —
recent observations get higher weight, older ones decay. It does not
require differencing or parameter search. The model notation `ETS(A,A,A)`
means additive error, additive trend, additive seasonality.

ETS is often more robust on shorter or noisier series and serves as a
useful benchmark against SARIMA.

```{r ets_fit, message=FALSE, warning=FALSE}
ets_models <- map(setNames(cities, cities), function(city) {
  ts_data <- build_ts(city)
  ets(ts_data)
})

map_dfr(cities, function(city) {
  m <- ets_models[[city]]
  tibble(
    City  = city,
    Model = m$method,
    AIC   = round(m$aic, 1),
    BIC   = round(m$bic, 1)
  )
}) %>% print()
```

### ETS Forecasts — All Cities

```{r ets_forecast, fig.width=12, fig.height=8, message=FALSE, warning=FALSE}
ets_forecasts <- map(setNames(cities, cities), function(city) {
  forecast(ets_models[[city]], h = h)
})

ets_plots <- map(cities, function(city) {
  fc <- ets_forecasts[[city]]
  color <- CITY_COLORS[city]

  autoplot(fc) +
    scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
    labs(title = city, x = NULL, y = "Home Value") +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(face = "bold", color = color))
})

do.call(grid.arrange, c(ets_plots, ncol = 2))
```

---

## VII. Model Comparison

### SARIMA vs ETS — Accuracy Metrics

Both models are evaluated on their in-sample fit using MAE (Mean
Absolute Error) and RMSE (Root Mean Squared Error). Lower is better.

```{r model_comparison}
comparison <- map_dfr(cities, function(city) {
  ts_data <- build_ts(city)

  sarima_acc <- accuracy(sarima_models[[city]])
  ets_acc    <- accuracy(ets_models[[city]])

  tibble(
    City       = city,
    SARIMA_MAE  = round(sarima_acc[, "MAE"], 0),
    SARIMA_RMSE = round(sarima_acc[, "RMSE"], 0),
    ETS_MAE     = round(ets_acc[, "MAE"], 0),
    ETS_RMSE    = round(ets_acc[, "RMSE"], 0),
    Better      = ifelse(sarima_acc[, "RMSE"] < ets_acc[, "RMSE"], "SARIMA", "ETS")
  )
})

print(comparison)
```

### 12-Month Forecast Summary

```{r forecast_summary}
forecast_summary <- map_dfr(cities, function(city) {
  sarima_fc <- sarima_forecasts[[city]]
  ets_fc    <- ets_forecasts[[city]]

  current <- df %>% filter(RegionName == city) %>% slice_tail(n = 1) %>% pull(home_value)

  tibble(
    City              = city,
    Current_Value     = round(current, 0),
    SARIMA_12mo       = round(as.numeric(tail(sarima_fc$mean, 1)), 0),
    ETS_12mo          = round(as.numeric(tail(ets_fc$mean, 1)), 0),
    SARIMA_Change_Pct = round((as.numeric(tail(sarima_fc$mean, 1)) / current - 1) * 100, 1),
    ETS_Change_Pct    = round((as.numeric(tail(ets_fc$mean, 1)) / current - 1) * 100, 1)
  )
})

print(forecast_summary)
```

---

## VIII. Conclusions

### Key Findings

**1. All five cities share the same structural pattern**
STL decomposition shows a dominant upward trend with modest seasonal
amplitude (~$5–10K peak-to-trough in Merced). Seasonality matters but
trend is the primary driver of long-run value.

**2. The market is coming off a post-COVID peak**
Year-over-year growth peaked at 20–25% in mid-2022 across all cities.
By early 2026 the YoY rate has returned to low single digits, consistent
with a normalization rather than a correction.

**3. Los Banos leads on absolute value despite smaller population**
The Bay Area commuter premium (documented in Part 3 of this series)
is visible in home values too — Los Banos consistently trades above
Merced and Atwater since ~2018.

**4. SARIMA and ETS produce similar 12-month forecasts**
Where both models agree the forecast is more reliable. Large
divergence between SARIMA and ETS signals higher uncertainty —
treat those forecasts with wider confidence intervals.

**5. Modest appreciation expected over the next 12 months**
Both models project low single-digit growth for Merced County cities,
consistent with the broader Central Valley cooling from 2022 peaks.
The GFC-era collapse (2007–2012) shows downside risk is real but
requires a significant macro shock to materialize.

---

### Series Summary

| Part | Project | Tool | Focus |
|---|---|---|---|
| 1 | Local Housing Market | Python | Los Banos deep dive |
| 2 | Merced County Housing | Python | 5-city comparative analysis |
| 3 | Local Rental Market | Python | Rent trends + price-to-rent |
| 4 | Housing Time Series | R (this project) | SARIMA/ETS 12-month forecast |

---

**Series:** Merced County Housing Analysis
**Author:** Jorge Reyes-Ornelas
Data Analyst | Wine Operations Specialist | MS Data Analytics Candidate
