---
title: "Wine Quality Statistical Analysis"
name: "Jorge RO"
date: 05/11/26
output: github_document
editor_options:
  markdown:
    wrap: 72
---

# The Chemistry of Quality

A Statistical Inquiry into What Makes Wine Good

## Project Overview

The wine industry spends billions on sensory evaluation — trained panels,
blind tastings, and expert scoring. But can chemistry alone predict quality?
This project applies classical statistical methods to the UCI Wine Quality
dataset to determine which physicochemical properties are statistically
significant predictors of wine quality, and whether red and white wines
obey different chemical rules.

Where the companion Python project (ML Wine Quality) asks *can a model predict
quality?*, this analysis asks *why* — using ANOVA, t-tests, correlation, and
linear regression to surface the structural chemical differences between
good and bad wines.

**Dataset:** 6,497 wines (1,599 red + 4,898 white), 11 chemical features,
quality scores 3–9.

---

## I. Environment Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(broom)
library(gridExtra)
library(scales)

# Load data
wine_df <- read_csv("winequality-combined.csv")

# Create binary quality label — matches Python project threshold
wine_df <- wine_df %>%
  mutate(quality_label = ifelse(quality >= 6, "Good", "Bad"),
         quality_label = factor(quality_label, levels = c("Bad", "Good")),
         wine_type = factor(wine_type))

cat("Dataset shape:", nrow(wine_df), "rows,", ncol(wine_df), "columns\n")
cat("Red wines:", sum(wine_df$wine_type == "red"), "\n")
cat("White wines:", sum(wine_df$wine_type == "white"), "\n")
cat("Good wines (>=6):", sum(wine_df$quality_label == "Good"), "\n")
cat("Bad wines (<6):", sum(wine_df$quality_label == "Bad"), "\n")
```

---

## II. Exploratory Overview

### Quality Score Distribution

```{r quality_dist, fig.width=10, fig.height=5}
p1 <- ggplot(wine_df, aes(x = factor(quality), fill = quality_label)) +
  geom_bar(alpha = 0.85, color = "white") +
  scale_fill_manual(values = c("Bad" = "#C0392B", "Good" = "#27AE60")) +
  labs(title = "Quality Score Distribution",
       subtitle = "Scores 3–9 | Threshold: Good ≥ 6",
       x = "Quality Score", y = "Count", fill = "Label") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(wine_df, aes(x = wine_type, fill = quality_label)) +
  geom_bar(position = "fill", alpha = 0.85, color = "white") +
  scale_fill_manual(values = c("Bad" = "#C0392B", "Good" = "#27AE60")) +
  scale_y_continuous(labels = percent) +
  labs(title = "Good vs. Bad Wine by Type",
       x = "Wine Type", y = "Proportion", fill = "Label") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

grid.arrange(p1, p2, ncol = 2)
```

---

## III. Feature Distributions — Good vs. Bad Wine

### Do chemical profiles differ between good and bad wines?

Before formal testing, we visualize the distribution of each feature
split by quality label. Features where the distributions separate clearly
are likely significant predictors.

```{r feature_distributions, fig.width=14, fig.height=10}
features <- c("alcohol", "volatile.acidity", "sulphates", "citric.acid",
              "density", "chlorides", "pH", "fixed.acidity",
              "residual.sugar", "free.sulfur.dioxide", "total.sulfur.dioxide")

# Rename columns for R compatibility
wine_plot <- wine_df %>%
  rename(
    volatile.acidity = `volatile acidity`,
    citric.acid = `citric acid`,
    residual.sugar = `residual sugar`,
    free.sulfur.dioxide = `free sulfur dioxide`,
    total.sulfur.dioxide = `total sulfur dioxide`,
    fixed.acidity = `fixed acidity`
  )

wine_long <- wine_plot %>%
  select(all_of(features), quality_label) %>%
  pivot_longer(-quality_label, names_to = "feature", values_to = "value")

ggplot(wine_long, aes(x = quality_label, y = value, fill = quality_label)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.5, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Bad" = "#C0392B", "Good" = "#27AE60")) +
  facet_wrap(~feature, scales = "free_y", ncol = 4) +
  labs(title = "Chemical Feature Distributions: Good vs. Bad Wine",
       subtitle = "Clear separation = strong predictor",
       x = NULL, y = "Value", fill = "Quality") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

---

## IV. Hypothesis Testing — Which Features Are Statistically Significant?

### Welch's Two-Sample t-Test (Good vs. Bad)

We run Welch's t-test for each chemical feature comparing good vs. bad
wines. Welch's variant does not assume equal variance — appropriate here
since group sizes are unequal (4,113 good vs. 2,384 bad).

**H₀:** Mean chemical value is equal between good and bad wines
**H₁:** Mean chemical value differs between good and bad wines

```{r ttests}
wine_test <- wine_df %>%
  rename(
    volatile_acidity = `volatile acidity`,
    citric_acid = `citric acid`,
    residual_sugar = `residual sugar`,
    free_sulfur_dioxide = `free sulfur dioxide`,
    total_sulfur_dioxide = `total sulfur dioxide`,
    fixed_acidity = `fixed acidity`
  )

feature_cols <- c("alcohol", "volatile_acidity", "sulphates", "citric_acid",
                  "density", "chlorides", "pH", "fixed_acidity",
                  "residual_sugar", "free_sulfur_dioxide", "total_sulfur_dioxide")

ttest_results <- map_dfr(feature_cols, function(feat) {
  good_vals <- wine_test %>% filter(quality_label == "Good") %>% pull(!!sym(feat))
  bad_vals  <- wine_test %>% filter(quality_label == "Bad")  %>% pull(!!sym(feat))

  test <- t.test(good_vals, bad_vals)

  tibble(
    Feature       = feat,
    Mean_Good     = mean(good_vals),
    Mean_Bad      = mean(bad_vals),
    Difference    = mean(good_vals) - mean(bad_vals),
    t_statistic   = test$statistic,
    p_value       = test$p.value,
    Significant   = ifelse(test$p.value < 0.05, "Yes ***", "No")
  )
}) %>%
  arrange(p_value)

print(ttest_results %>% mutate(across(where(is.numeric), ~round(., 4))),
      n = Inf)
```

### Visualizing Effect Sizes

```{r effect_size_plot, fig.width=10, fig.height=6}
ttest_results %>%
  mutate(Feature = fct_reorder(Feature, abs(t_statistic)),
         Direction = ifelse(Difference > 0, "Higher in Good", "Lower in Good")) %>%
  ggplot(aes(x = Feature, y = t_statistic, fill = Direction)) +
  geom_col(alpha = 0.85, color = "white") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("Higher in Good" = "#27AE60", "Lower in Good" = "#C0392B")) +
  coord_flip() +
  labs(title = "t-Statistics by Feature: Good vs. Bad Wine",
       subtitle = "Dashed lines = ±2 (approximate significance threshold)",
       x = NULL, y = "t-Statistic", fill = "Direction") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## V. ANOVA — Does Quality Score Differ by Wine Type?

### Do red and white wines achieve quality through different chemistry?

```{r anova}
# One-way ANOVA: quality score ~ wine type
anova_model <- aov(quality ~ wine_type, data = wine_df)
cat("ANOVA: Quality Score ~ Wine Type\n")
print(summary(anova_model))
```
```{r anova_summary}
# Mean quality by type
wine_df %>%
  group_by(wine_type) %>%
  summarise(
    n = n(),
    mean_quality = round(mean(quality), 3),
    sd_quality   = round(sd(quality), 3),
    pct_good     = round(mean(quality_label == "Good") * 100, 1)
  ) %>%
  print()
```

### Key Chemical Differences Between Red and White Wines

```{r red_vs_white, fig.width=12, fig.height=8}
wine_test %>%
  select(wine_type, all_of(feature_cols)) %>%
  pivot_longer(-wine_type, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = wine_type, y = value, fill = wine_type)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.4, outlier.alpha = 0.2) +
  scale_fill_manual(values = c("red" = "#8B0000", "white" = "#C8A2C8")) +
  facet_wrap(~feature, scales = "free_y", ncol = 4) +
  labs(title = "Chemical Profiles: Red vs. White Wine",
       subtitle = "Structural chemical differences between wine types",
       x = NULL, y = "Value", fill = "Type") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

---

## VI. Correlation Analysis

### How do features relate to each other and to quality?

```{r correlation, fig.width=10, fig.height=9}
cor_data <- wine_test %>%
  select(all_of(feature_cols), quality) %>%
  cor(use = "complete.obs")

corrplot(cor_data,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.srt = 45,
         tl.cex = 0.85,
         addCoef.col = "black",
         number.cex = 0.65,
         col = colorRampPalette(c("#C0392B", "white", "#27AE60"))(200),
         title = "Feature Correlation Matrix (incl. Quality)",
         mar = c(0, 0, 2, 0))
```

---

## VII. Linear Regression — Predicting Quality Score

### Which features are the strongest linear predictors of quality?

```{r regression}
lm_model <- lm(quality ~ alcohol + volatile_acidity + sulphates +
                 citric_acid + density + chlorides + pH +
                 fixed_acidity + residual_sugar +
                 free_sulfur_dioxide + total_sulfur_dioxide + wine_type,
               data = wine_test)

cat("Linear Regression: Quality ~ All Features\n\n")
print(summary(lm_model))
```

### Coefficient Plot — Direction and Significance of Each Predictor

```{r coef_plot, fig.width=10, fig.height=7}
coef_data <- tidy(lm_model, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

density_coef <- coef_data %>% filter(term == "density") %>% pull(estimate) %>% round(1)

coef_data %>%
  filter(term != "density") %>%
  mutate(term = fct_reorder(term, estimate),
         significant = ifelse(p.value < 0.05, "p < 0.05", "p ≥ 0.05")) %>%
  ggplot(aes(x = estimate, y = term, color = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("p < 0.05" = "#1f77b4", "p ≥ 0.05" = "#aaaaaa")) +
  labs(title = "Regression Coefficients: Predictors of Wine Quality",
       subtitle = paste0("Points = estimate | Bars = 95% CI | density excluded (coef ≈ ",
                         density_coef, "; large due to unit scale)"),
       x = "Coefficient Estimate", y = NULL, color = "Significance") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## VIII. Conclusions

### Summary of Statistical Findings

```{r summary_table}
ttest_results %>%
  select(Feature, Mean_Good, Mean_Bad, Difference, p_value, Significant) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  arrange(p_value) %>%
  knitr::kable(caption = "t-Test Results: Good vs. Bad Wine (all features)")
```

### Key Findings

**1. Alcohol is the strongest positive predictor of quality**
Good wines average significantly higher alcohol content. The t-test
confirms this is the most statistically significant single feature
(highest |t-statistic|). Higher alcohol correlates with riper grapes
and more complete fermentation — signals of quality viticulture.

**2. Volatile acidity is the strongest negative predictor**
High volatile acidity (acetic acid) is the clearest chemical marker of
bad wine. The mean difference between good and bad wines is substantial
and highly significant. Acetic acid produces a vinegar-like taste that
directly degrades sensory quality.

**3. Sulphates and density round out the top predictors**
Good wines have higher sulphate levels (a preservative that contributes
to SO₂ levels and wine stability) and lower density (correlates with
higher alcohol and drier style).

**4. Red and white wines have structurally different chemistry**
ANOVA confirms that wine type is a significant predictor of quality score.
The faceted box plots show red and white wines differ substantially in
volatile acidity, residual sugar, free SO₂, and total SO₂ — suggesting
quality standards are chemically distinct between the two types.

**5. Linear regression explains ~30% of quality variance**
The OLS model with all 11 features achieves an R² of approximately 0.30.
This aligns with the broader finding that chemistry alone captures a
meaningful but incomplete picture of quality — sensory factors,
winemaking technique, and evaluator subjectivity account for the rest.

---

### Relationship to ML Wine Quality Project

This statistical analysis complements the companion Python ML project:

| Approach | Tool | Question | Answer |
|---|---|---|---|
| Statistical | R (this project) | *Why* are wines good or bad? | Alcohol ↑, volatile acidity ↓ |
| Machine Learning | Python (ML project) | *Can we predict* quality? | Yes — 83.8% accuracy (RF) |

The statistical approach surfaces interpretable, causal-adjacent insights.
The ML approach maximizes predictive power. Together they tell a complete story.

---

**Series:** Wine Analytics
**Part 1:** Wine Quality Statistical Analysis — R (this project)
**Part 2:** ML Wine Quality Predictor — Python + Streamlit

**Author:** Jorge Reyes-Ornelas
Data Analyst | Wine Operations Specialist | MS Data Analytics Candidate
