---
title: "California English Learner Data — Exploratory Analysis"
author: "Jorge Reyes-Ornelas"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
library(tidyverse)
library(scales)
library(knitr)
```

# Data Sources

Three school-level files from the California Department of Education (CDE):

| File | Description | Grain |
|------|-------------|-------|
| `fileselsch_2026.txt` | English Learner (EL) counts — 2025–26 | School × Language × Grade |
| `filesfepsch_2026.txt` | Fluent English Proficient (FEP) counts — 2025–26 | School × Language × Grade |
| `filesreclass.txt` | Reclassified Fluent English counts — last available (2019–20) | School (aggregate) |

---

# 1. Load Data

```{r load}
el  <- read_tsv("data/fileselsch_2026.txt",  show_col_types = FALSE)
fep <- read_tsv("data/filesfepsch_2026.txt", show_col_types = FALSE)
rc  <- read_tsv("data/filesreclass.txt",     show_col_types = FALSE)
```

---

# 2. Structure & Dimensions

```{r structure}
cat("=== EL ===\n");  glimpse(el)
cat("\n=== FEP ===\n"); glimpse(fep)
cat("\n=== RECLASS ===\n"); glimpse(rc)
```

```{r dims}
tibble(
  File    = c("EL", "FEP", "Reclass"),
  Rows    = c(nrow(el), nrow(fep), nrow(rc)),
  Columns = c(ncol(el), ncol(fep), ncol(rc))
) |> kable(caption = "Dataset dimensions")
```

---

# 3. Missing Values

```{r missing}
missing_summary <- function(df, name) {
  df |>
    summarise(across(everything(), ~ sum(is.na(.)))) |>
    pivot_longer(everything(), names_to = "column", values_to = "na_count") |>
    filter(na_count > 0) |>
    mutate(file = name)
}

bind_rows(
  missing_summary(el,  "EL"),
  missing_summary(fep, "FEP"),
  missing_summary(rc,  "Reclass")
) |>
  select(file, column, na_count) |>
  kable(caption = "Columns with missing values")
```

---

# 4. Top Languages — EL

```{r top-langs-el}
el |>
  group_by(LANGUAGE) |>
  summarise(total = sum(TOTAL_EL, na.rm = TRUE)) |>
  arrange(desc(total)) |>
  slice_head(n = 20) |>
  ggplot(aes(x = reorder(LANGUAGE, total), y = total)) +
  geom_col(fill = "#2c7bb6") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Top 20 Languages — English Learners (Statewide)",
       x = NULL, y = "Total EL Students") +
  theme_minimal()
```

```{r top-langs-el-table}
el |>
  group_by(LC, LANGUAGE) |>
  summarise(total_el = sum(TOTAL_EL, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_el)) |>
  slice_head(n = 20) |>
  mutate(pct = percent(total_el / sum(el$TOTAL_EL, na.rm = TRUE), accuracy = 0.1)) |>
  kable(caption = "Top 20 EL Languages with statewide share")
```

---

# 5. Top Languages — FEP

```{r top-langs-fep}
fep |>
  group_by(LANGUAGE) |>
  summarise(total = sum(TOTAL, na.rm = TRUE)) |>
  arrange(desc(total)) |>
  slice_head(n = 20) |>
  ggplot(aes(x = reorder(LANGUAGE, total), y = total)) +
  geom_col(fill = "#1a9641") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Top 20 Languages — Fluent English Proficient (Statewide)",
       x = NULL, y = "Total FEP Students") +
  theme_minimal()
```

---

# 6. EL Grade Distribution (Statewide)

```{r grade-dist-el}
grade_cols <- c("GR_TK","GR_KN","GR_1","GR_2","GR_3","GR_4","GR_5",
                "GR_6","GR_7","GR_8","GR_9","GR_10","GR_11","GR_12","UNGR")

el |>
  summarise(across(all_of(grade_cols), ~ sum(., na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "grade", values_to = "count") |>
  mutate(grade = factor(grade, levels = grade_cols)) |>
  ggplot(aes(x = grade, y = count)) +
  geom_col(fill = "#2c7bb6") +
  scale_y_continuous(labels = comma) +
  labs(title = "EL Students by Grade Level (Statewide)",
       x = "Grade", y = "Students") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 7. FEP Grade Distribution (Statewide)

```{r grade-dist-fep}
fep |>
  summarise(across(all_of(grade_cols), ~ sum(., na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "grade", values_to = "count") |>
  mutate(grade = factor(grade, levels = grade_cols)) |>
  ggplot(aes(x = grade, y = count)) +
  geom_col(fill = "#1a9641") +
  scale_y_continuous(labels = comma) +
  labs(title = "FEP Students by Grade Level (Statewide)",
       x = "Grade", y = "Students") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 8. County Summaries

```{r county-el}
el |>
  group_by(COUNTY) |>
  summarise(total_el = sum(TOTAL_EL, na.rm = TRUE)) |>
  arrange(desc(total_el)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(COUNTY, total_el), y = total_el)) +
  geom_col(fill = "#d7191c") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Top 15 Counties by EL Enrollment", x = NULL, y = "Total EL Students") +
  theme_minimal()
```

---

# 9. Reclassification Overview

```{r reclass-overview}
rc |>
  summarise(
    total_el    = sum(EL,     na.rm = TRUE),
    total_fep   = sum(FEP,    na.rm = TRUE),
    total_reclass = sum(Reclass, na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "category", values_to = "count") |>
  mutate(
    category = recode(category,
      total_el = "English Learner",
      total_fep = "Fluent English Proficient",
      total_reclass = "Reclassified"
    )
  ) |>
  kable(caption = "Statewide EL / FEP / Reclassification counts (from Reclass file)")
```

```{r reclass-rate}
rc |>
  filter(EL > 0) |>
  mutate(reclass_rate = Reclass / EL) |>
  filter(is.finite(reclass_rate)) |>
  ggplot(aes(x = reclass_rate)) +
  geom_histogram(bins = 60, fill = "#fdae61", color = "white") +
  scale_x_continuous(labels = percent_format()) +
  labs(title = "Distribution of Reclassification Rates by School",
       subtitle = "Reclass / EL enrollment, schools with EL > 0",
       x = "Reclassification Rate", y = "Schools") +
  theme_minimal()
```

```{r reclass-county}
rc |>
  group_by(County) |>
  summarise(
    el      = sum(EL,      na.rm = TRUE),
    reclass = sum(Reclass, na.rm = TRUE),
    rate    = reclass / el
  ) |>
  filter(el > 0) |>
  arrange(desc(el)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(County, rate), y = rate)) +
  geom_col(fill = "#fdae61") +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(title = "Reclassification Rate — Top 15 Counties by EL Enrollment",
       x = NULL, y = "Reclassification Rate") +
  theme_minimal()
```

---

# 10. Join Preview (for next project)

The three files share the `CDS` school code. Reclass uses title-case column names; EL and FEP use uppercase. A full join will need:

- Rename `County / District / School` in Reclass to `COUNTY / DISTRICT / SCHOOL`
- Pivot EL and FEP from wide (grade columns) to long before joining, or join at the school-aggregate level
- `TOTAL_EL` (EL file) vs `TOTAL` (FEP file) — standardize before binding

```{r join-preview}
# School-level aggregate from EL file
el_agg <- el |>
  group_by(CDS, COUNTY, DISTRICT, SCHOOL) |>
  summarise(total_el = sum(TOTAL_EL, na.rm = TRUE), .groups = "drop")

# School-level aggregate from FEP file
fep_agg <- fep |>
  group_by(CDS, COUNTY, DISTRICT, SCHOOL) |>
  summarise(total_fep = sum(TOTAL, na.rm = TRUE), .groups = "drop")

# Reclass — standardize column names
rc_clean <- rc |>
  rename(COUNTY = County, DISTRICT = District, SCHOOL = School)

# Preview join
el_agg |>
  left_join(fep_agg, by = c("CDS","COUNTY","DISTRICT","SCHOOL")) |>
  left_join(rc_clean, by = c("CDS","COUNTY","DISTRICT","SCHOOL")) |>
  slice_head(n = 10) |>
  kable(caption = "Preview: school-level join of EL + FEP + Reclass")
```

---

# Session Info

```{r session}
sessionInfo()
```
