---
title: "Central Valley Crop Economics"
author: "Jorge Reyes-Ornelas"
date: "`r Sys.Date()`"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    theme: flatly
    source_code: embed
---

```{css, echo=FALSE}
/* Force white text in value boxes — flatly theme renders near-invisible otherwise */
.value-box .value,
.value-box .caption,
.value-box .inner h3,
.value-box p,
.value-box-body,
.value-box-body .value,
.value-box-body p {
  color: #ffffff !important;
  opacity: 1 !important;
}
```

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(scales)
library(knitr)
library(RColorBrewer)

PROJ_ROOT <- path.expand("~/Documents/GitHub/Portfolio/R_Projects/Central_Valley_Ag_Economics")
clean     <- readRDS(file.path(PROJ_ROOT, "data/cv_ag_clean.rds"))
derived   <- readRDS(file.path(PROJ_ROOT, "data/cv_ag_derived.rds"))
yield_idx <- readRDS(file.path(PROJ_ROOT, "data/cv_ag_yield_index.rds"))

COUNTIES    <- c("Fresno","Tulare","Kings","Merced","San Joaquin","Stanislaus","Madera")
COUNTY_COLS <- setNames(brewer.pal(7, "Set1"), COUNTIES)
CROP_COLS   <- setNames(brewer.pal(6, "Set2"),
                        c("Almonds","Cotton","Grapes","Milk","Tomatoes","Walnuts"))

# Separate county vs statewide for charts that need each
clean_county <- clean |> filter(geo_level == "county")
clean_state  <- clean |> filter(geo_level == "state")

latest_year <- clean_state |>
  filter(metric_label == "Value of Production") |>
  pull(year) |> max(na.rm = TRUE)

if (!is.finite(latest_year)) latest_year <- max(clean$year, na.rm = TRUE)

# County value data is Census-only (2012/2017/2022) — use its own latest year
latest_year_county <- clean_county |>
  filter(metric_label == "Value of Production") |>
  pull(year) |> max(na.rm = TRUE)

if (!is.finite(latest_year_county)) latest_year_county <- latest_year

# Helper: compute paper-coordinate annotations for subplot titles.
# Plotly does NOT remap yref="paper" y=1 from individual subplots after subplot().
subplot_titles <- function(labels, nrows, margin = 0.07) {
  n     <- length(labels)
  ncols <- ceiling(n / nrows)
  sw    <- (1 - (ncols - 1) * margin) / ncols
  sh    <- (1 - (nrows - 1) * margin) / nrows
  lapply(seq_along(labels), function(i) {
    r <- ceiling(i / ncols)
    c <- ((i - 1) %% ncols) + 1
    list(
      text      = labels[[i]],
      x         = (c - 1) * (sw + margin) + sw / 2,
      xref      = "paper",
      y         = 1 - (r - 1) * (sh + margin),
      yref      = "paper",
      showarrow = FALSE,
      xanchor   = "center",
      yanchor   = "bottom",
      font      = list(size = 11, color = "#333")
    )
  })
}
```

Overview
=====================================

Row
-------------------------------------

### Total Value of Production — `r latest_year` {.value-box}
```{r}
total_val <- clean_state |>
  filter(year == latest_year, metric_label == "Value of Production") |>
  summarise(total = sum(value, na.rm = TRUE)) |> pull(total)

val_label <- if (length(total_val) == 0 || total_val == 0) "N/A" else paste0("$", round(total_val / 1e9, 1), "B")
valueBox(val_label, icon = "fa-seedling", color = "success")
```

### Counties Tracked {.value-box}
```{r}
valueBox(length(COUNTIES), icon = "fa-map-marker", color = "primary")
```

### Crops Tracked {.value-box}
```{r}
valueBox(clean |> distinct(crop_label) |> nrow(), icon = "fa-leaf", color = "warning")
```

Row
-------------------------------------

### Value of Production by County & Crop — `r latest_year_county` (Census)
```{r}
overview_df <- clean_county |>
  filter(year == latest_year_county, metric_label == "Value of Production") |>
  group_by(county, crop_label) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") |>
  mutate(tip = paste0(county, "<br>", crop_label, "<br>$", round(value/1e6,1), "M"))

county_order <- overview_df |>
  group_by(county) |> summarise(tot = sum(value)) |>
  arrange(tot) |> pull(county)

plot_ly(overview_df,
        x = ~value / 1e6,
        y = ~factor(county, levels = county_order),
        color = ~crop_label,
        colors = CROP_COLS,
        type = "bar", orientation = "h",
        text = ~tip, hoverinfo = "text") |>
  layout(barmode = "stack",
         xaxis = list(title = "Value ($ Millions)", tickprefix = "$"),
         yaxis = list(title = ""),
         legend = list(title = list(text = "Crop")))
```

### Top Crops by Statewide Value — `r latest_year`
```{r}
top_crops <- clean_state |>
  filter(year == latest_year, metric_label == "Value of Production") |>
  group_by(crop_label) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
  arrange(total) |>
  mutate(tip = paste0(crop_label, ": $", round(total/1e6,1), "M"))

plot_ly(top_crops,
        x = ~total / 1e6,
        y = ~factor(crop_label, levels = top_crops$crop_label),
        color = ~crop_label, colors = CROP_COLS,
        type = "bar", orientation = "h",
        text = ~tip, hoverinfo = "text",
        showlegend = FALSE) |>
  layout(xaxis = list(title = "Value ($ Millions)", tickprefix = "$"),
         yaxis = list(title = ""))
```

Crop Trends
=====================================

Row
-------------------------------------

### Value of Production Over Time — All Counties
```{r}
trend_df <- clean_state |>
  filter(metric_label == "Value of Production") |>
  group_by(year, crop_label) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
  mutate(tip = paste0(crop_label, " (", year, "): $", round(total/1e6,1), "M"))

plot_ly(trend_df, x = ~year, y = ~total/1e6, color = ~crop_label,
        colors = CROP_COLS, type = "scatter", mode = "lines+markers",
        text = ~tip, hoverinfo = "text") |>
  layout(xaxis = list(title = "Year"),
         yaxis = list(title = "Value ($ Millions)", tickprefix = "$"),
         legend = list(title = list(text = "Crop")))
```

Row
-------------------------------------

### Acres Harvested Over Time
```{r}
acres_df <- clean |>
  filter(metric_label == "Acres Harvested") |>
  group_by(year, crop_label) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
  mutate(tip = paste0(crop_label, " (", year, "): ", round(total/1e3,1), "K acres"))

plot_ly(acres_df, x = ~year, y = ~total/1e3, color = ~crop_label,
        colors = CROP_COLS, type = "scatter", mode = "lines+markers",
        text = ~tip, hoverinfo = "text") |>
  layout(xaxis = list(title = "Year"),
         yaxis = list(title = "Thousand Acres"),
         legend = list(title = list(text = "Crop")))
```

Price & Yield
=====================================

Row
-------------------------------------

### Price Received Over Time
```{r}
price_df <- clean_state |>
  filter(metric_label == "Price Received") |>
  group_by(year, crop_label) |>
  summarise(avg_price = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(tip = paste0(crop_label, " (", year, "): $", round(avg_price, 2)))

plot_ly(price_df, x = ~year, y = ~avg_price, color = ~crop_label,
        colors = CROP_COLS, type = "scatter", mode = "lines+markers",
        text = ~tip, hoverinfo = "text") |>
  layout(xaxis = list(title = "Year"),
         yaxis = list(title = "Avg Price Received", tickprefix = "$"),
         legend = list(title = list(text = "Crop")))
```

### Value per Acre by Crop
```{r}
if (nrow(derived) == 0) {
  plotly_empty() |> layout(title = list(text = "No value-per-acre data (requires both Acres Harvested and Value of Production)"))
} else {
  derived_plot <- derived |>
    mutate(tip = paste0(crop_label, " — ", county,
                        " (", year, "): $", round(value_per_acre)))

  plot_ly(derived_plot, x = ~year, y = ~value_per_acre,
          split = ~paste(crop_label, county),
          color = ~crop_label, colors = CROP_COLS,
          type = "scatter", mode = "lines",
          opacity = 0.6,
          text = ~tip, hoverinfo = "text") |>
    layout(xaxis = list(title = "Year"),
           yaxis = list(title = "Value per Acre ($)", tickprefix = "$"),
           legend = list(title = list(text = "Crop")))
}
```

County Comparison
=====================================

Row
-------------------------------------

### Value of Production by County — by Crop
```{r}
county_df <- clean_county |>
  filter(metric_label == "Value of Production", county %in% COUNTIES) |>
  group_by(year, county, crop_label) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") |>
  mutate(tip = paste0(county, " (", year, "): $", round(value/1e6,1), "M"))

crops <- sort(unique(county_df$crop_label))

plots <- lapply(seq_along(crops), function(i) {
  cr  <- crops[i]
  d   <- filter(county_df, crop_label == cr)
  show_legend <- (i == 1)
  plot_ly(d, x = ~year, y = ~value/1e6, color = ~county,
          colors = COUNTY_COLS, type = "scatter", mode = "lines",
          text = ~tip, hoverinfo = "text",
          showlegend = show_legend, legendgroup = ~county) |>
    layout(yaxis = list(tickprefix = "$", title = "$M"))
})

if (length(plots) == 0) {
  plotly_empty() |> layout(title = list(text = "No Value of Production data available"))
} else {
  nr <- min(3, length(plots))
  subplot(plots, nrows = nr, shareX = FALSE, shareY = FALSE,
          titleX = FALSE, titleY = FALSE, margin = 0.07) |>
    layout(annotations = subplot_titles(crops, nrows = nr, margin = 0.07))
}
```

COVID & Climate Impact
=====================================

Row
-------------------------------------

### Year-over-Year Change in Value of Production
```{r}
yoy_df <- clean_state |>
  filter(metric_label == "Value of Production") |>
  group_by(year, crop_label) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") |>
  arrange(crop_label, year) |>
  group_by(crop_label) |>
  mutate(pct_change = (value - lag(value)) / lag(value)) |>
  filter(!is.na(pct_change)) |>
  ungroup() |>
  mutate(
    color_val = if_else(pct_change > 0, "#1a9641", "#d7191c"),
    tip = paste0(crop_label, " (", year, "): ", round(pct_change*100,1), "%")
  )

crops_yoy <- sort(unique(yoy_df$crop_label))
x_min <- min(yoy_df$year); x_max <- max(yoy_df$year)

plots_yoy <- lapply(seq_along(crops_yoy), function(i) {
  cr <- crops_yoy[i]
  d  <- filter(yoy_df, crop_label == cr)
  plot_ly(d, x = ~year, y = ~pct_change,
          type = "bar", marker = list(color = d$color_val),
          text = ~tip, hoverinfo = "text",
          showlegend = FALSE) |>
    layout(
      shapes = list(list(
        type = "rect", x0 = 2019.5, x1 = 2022.5,
        y0 = min(d$pct_change, na.rm=TRUE),
        y1 = max(d$pct_change, na.rm=TRUE),
        fillcolor = "rgba(215,25,28,0.08)", line = list(width = 0)
      )),
      yaxis = list(tickformat = ".0%", title = "")
    )
})

if (length(plots_yoy) == 0) {
  plotly_empty() |> layout(title = list(text = "No Value of Production data available"))
} else {
  nr <- min(2, length(plots_yoy))
  subplot(plots_yoy, nrows = nr, shareX = FALSE, shareY = FALSE,
          titleX = FALSE, titleY = FALSE, margin = 0.08) |>
    layout(annotations = subplot_titles(crops_yoy, nrows = nr, margin = 0.08))
}
```

Row
-------------------------------------

### Yield Index (2015 = 100)
```{r}
yield_plot <- yield_idx |>
  filter(!is.na(yield_index), county %in% COUNTIES) |>
  mutate(tip = paste0(county, " (", year, "): ", round(yield_index, 1)))

crops_yi <- sort(unique(yield_plot$crop_label))

plots_yi <- lapply(seq_along(crops_yi), function(i) {
  cr <- crops_yi[i]
  d  <- filter(yield_plot, crop_label == cr)
  show_legend <- (i == 1)
  plot_ly(d, x = ~year, y = ~yield_index, color = ~county,
          colors = COUNTY_COLS, type = "scatter", mode = "lines",
          text = ~tip, hoverinfo = "text",
          showlegend = show_legend, legendgroup = ~county) |>
    layout(
      shapes = list(list(
        type = "line", x0 = min(d$year), x1 = max(d$year),
        y0 = 100, y1 = 100,
        line = list(dash = "dash", color = "gray", width = 1)
      )),
      yaxis = list(title = "Index (2015=100)")
    )
})

if (length(plots_yi) == 0) {
  plotly_empty() |> layout(title = list(text = "No yield index data available"))
} else {
  nr <- min(3, length(plots_yi))
  subplot(plots_yi, nrows = nr, shareX = FALSE, shareY = FALSE,
          titleX = FALSE, titleY = FALSE, margin = 0.07) |>
    layout(annotations = subplot_titles(crops_yi, nrows = nr, margin = 0.07))
}
```

Key Findings
=====================================

Row
-------------------------------------

### Analysis

**Central Valley Crop Economics: 2010–`r latest_year`**

The seven counties in this analysis — Fresno, Tulare, Kings, Merced, San Joaquin,
Stanislaus, and Madera — form the agricultural backbone of California and one of
the most productive farming regions on Earth. Together they produce the majority
of the state's almonds, walnuts, grapes, processing tomatoes, and dairy output.
This analysis examines how crop economics have shifted over fifteen years, with
particular attention to structural disruptions from drought, COVID-19, and
commodity price cycles.

---

**1. Almonds dominate value per acre — but at a cost.**
Almond acreage expanded dramatically through the 2010s as growers chased high
commodity prices. By the mid-2010s, however, global oversupply began compressing
prices. The value-per-acre trends reveal the boom-and-contraction cycle clearly:
rapid expansion followed by price erosion that squeezed margins even as acreage
held steady. This is a textbook commodity cycle — overinvestment chasing a price
signal that no longer holds by the time new orchards mature.

**2. Drought years (2012–2016, 2020–2022) appear in the yield index.**
California's multi-year droughts are legible in the yield data for water-intensive
crops. Counties on the west side of the San Joaquin Valley — Kings and Fresno
in particular — show steeper yield index declines during drought windows, reflecting
their greater dependence on groundwater and their distance from Sierra Nevada
snowpack runoff. The Sustainable Groundwater Management Act (SGMA, 2014) is
beginning to constrain pumping, which will reshape where and what can be grown
in the coming decade.

**3. COVID-19 disrupted price received more than production volume.**
The 2020–2021 window shows in the YoY price charts more than in the acreage or
yield data. Supply chains for labor-intensive crops — grapes and tomatoes especially
— faced harvest disruptions from farmworker illness and H-2A visa processing delays.
Processing tomato growers absorbed some shock through contract pricing, while fresh
grape markets faced steeper volatility. The overall value of production recovered
faster than prices, suggesting volume held up while margins compressed.

**4. Merced County is a dairy county first.**
Milk production dominates Merced's agricultural value in most years, reflecting
the county's role in California's Central Valley dairy belt. Row crops and grapes
play a secondary role. This matters for county-level economic resilience: dairy
is subject to federal milk marketing orders and national commodity price swings
in ways that tree crops are not, making Merced's ag economy more exposed to
federal policy than counties more diversified into nuts and grapes.

**5. The coming decade will be shaped by water, not weather.**
SGMA implementation, declining snowpack, and competition between agricultural and
municipal water users will force land conversion across the Valley. Crops requiring
deep, sustained irrigation — cotton, some row crops — will retreat. Tree crops
already in the ground will be managed defensively. The acreage trends in this
dataset already show early signs of this transition, and the next five years of
NASS data will make the structural shift visible.
